---
title: "lncRNA report"
subtitle: "norm_group report"
format:
  html:
    self-contained: true
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
    page-layout: full
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
params:
  contrast: ""
  group: ""
  de_summary: null
  cis_pairs: null
  cis_pairs_sig: null
  cis_window: 100000
  ora: null
  gsea: null
  heterogeneity: null
  heterogeneity_cv: null
  heterogeneity_concordant_frac: 0.50
  heterogeneity_discordant_frac: 0.25
  gene_catalog: null
  tpm: null
  species: ""
  genome_build: ""
  conservation_metric: "phastCons"
  conservation_configured: false
  synteny_configured: false
  synteny_target_species: ""
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(ggrepel)
library(ggridges)

contrast <- params$contrast
group    <- params$group

ORIGIN_COLORS <- c(
  "novel"                = "#c2185b",
  "annotated_lncrna"     = "#2e7d32",
  "reference_other"      = "#7f7f7f"
)

ORIGIN_LABELS <- c(
  "novel"                = "Novel lncRNAs",
  "annotated_lncrna"     = "Annotated lncRNAs (reference)",
  "reference_other"      = "Other reference genes (background)"
)

CLASS_CODE_LEGEND <- c(
  "u" = "Intergenic (lincRNA-like)",
  "x" = "Antisense",
  "i" = "Intronic",
  "j" = "Novel isoform (multi-exon, at least one splice site shared)",
  "o" = "Overlap (same strand, other exonic overlap)",
  "c" = "Contained (within reference intron)",
  "k" = "Contained (within reference intron, opposite strand)",
  "m" = "Retained intron (same strand)",
  "n" = "Retained intron (opposite strand)",
  "s" = "Intron overlap (opposite strand)",
  "y" = "Contained within reference pseudogene",
  "p" = "Polymerase run-on",
  "r" = "Repeat",
  "e" = "Exonic overlap (opposite strand, single exon)"
)

format_class_code <- function(cc) {
  if (is.na(cc) || cc == "") return("unknown")
  parts <- strsplit(cc, ",\\s*")[[1]]
  parts <- parts[nzchar(parts)]
  desc <- CLASS_CODE_LEGEND[parts]
  desc[is.na(desc)] <- parts[is.na(desc)]
  paste0(sprintf("%s (%s)", parts, desc), collapse = "; ")
}

# Truncate long gene-list columns with click-to-expand via <details>
compact_gene_col <- function(x, max_show = 5) {
  vapply(x, function(s) {
    if (is.na(s) || nchar(s) == 0) return(NA_character_)
    sep <- if (grepl("/", s, fixed = TRUE)) "/" else ",\\s*"
    genes <- strsplit(s, sep, perl = TRUE)[[1]]
    genes <- trimws(genes)
    genes <- genes[nzchar(genes)]
    n <- length(genes)
    if (n == 0) return("")
    shown <- paste(genes[seq_len(min(n, max_show))], collapse = ", ")
    full  <- paste(genes, collapse = ", ")
    if (n > max_show) {
      sprintf('<details><summary>%s <em>(+%d more)</em></summary>%s</details>',
              htmltools::htmlEscape(shown), n - max_show, htmltools::htmlEscape(full))
    } else {
      htmltools::htmlEscape(full)
    }
  }, character(1))
}

compact_list_columns <- function(tab) {
  for (nm in names(tab)) {
    x <- tab[[nm]]
    if (!is.character(x)) next
    if (any(nchar(x[!is.na(x)]) > 40)) {
      tab[[nm]] <- compact_gene_col(x)
    } else {
      tab[[nm]] <- htmltools::htmlEscape(x)
    }
  }
  tab
}
```

## Contrast: `r contrast`

```{r load, include=FALSE}
de <- tryCatch(
  read_tsv(params$de_summary, show_col_types = FALSE),
  error = function(e) NULL
)
de_ct <- if (!is.null(de) && "contrast" %in% colnames(de)) {
  filter(de, contrast == !!contrast)
} else { de }

catalog <- tryCatch(read_tsv(params$gene_catalog, show_col_types = FALSE),
                    error = function(e) NULL)

de_ann <- de_ct
if (!is.null(catalog) && nrow(catalog) > 0 && "gene_id" %in% colnames(de_ct)) {
  de_ann <- de_ct %>%
    left_join(select(catalog, gene_id, gene_name, gene_biotype, origin, class_code,
                     conservation_mean, conservation_pct, syntenic_locus, syntenic_target_gene_id),
              by = "gene_id")
}

# Conservation/synteny gating: a section is only shown when the configured
# module produced data (not merely when the column exists), otherwise an
# all-NA column would render empty/confusing plots. Columns are always present
# in the catalog (NA when the module was skipped).
has_conservation <- isTRUE(params$conservation_configured) &&
  !is.null(catalog) && "conservation_mean" %in% colnames(catalog) &&
  any(!is.na(catalog$conservation_mean))
has_synteny <- isTRUE(params$synteny_configured) &&
  !is.null(catalog) && "syntenic_locus" %in% colnames(catalog) &&
  any(!is.na(catalog$syntenic_locus))

cis <- tryCatch(read_tsv(params$cis_pairs, show_col_types = FALSE),
                error = function(e) NULL)
cis_sig <- tryCatch(read_tsv(params$cis_pairs_sig, show_col_types = FALSE),
                    error = function(e) NULL)

# Annotate cis pairs with DE stats from THIS contrast only
# (left_join keeps all cis rows — lncRNAs without a DE match get NA)
if (!is.null(cis_sig) && nrow(cis_sig) > 0 && !is.null(de_ct) && nrow(de_ct) > 0) {
  cis_sig <- cis_sig %>%
    left_join(
      select(de_ct, gene_id, lncrna_DE_log2FC = log2FoldChange, lncrna_DE_padj = padj),
      by = c("lncrna_id" = "gene_id")
    )
}

ora <- tryCatch(read_tsv(params$ora, show_col_types = FALSE),
                error = function(e) NULL)
gsea <- tryCatch(read_tsv(params$gsea, show_col_types = FALSE),
                 error = function(e) NULL)

# Restrict enrichment tables to the current contrast
if (!is.null(ora) && "contrast" %in% colnames(ora)) {
  ora <- filter(ora, contrast == !!contrast)
}
if (!is.null(gsea) && "contrast" %in% colnames(gsea)) {
  gsea <- filter(gsea, contrast == !!contrast)
}

# Heterogeneity diagnostic (optional feature, off by default): the group-level table is
# a header-only, 0-row placeholder when params.heterogeneity_analysis is disabled for
# this norm_group, so het_ct/het_cv_ct simply end up empty and the section is skipped.
het <- tryCatch(read_tsv(params$heterogeneity, show_col_types = FALSE),
                error = function(e) NULL)
het_cv <- tryCatch(read_tsv(params$heterogeneity_cv, show_col_types = FALSE),
                   error = function(e) NULL)
het_ct <- if (!is.null(het) && "contrast" %in% colnames(het)) {
  filter(het, contrast == !!contrast)
} else { NULL }
het_cv_ct <- if (!is.null(het_cv) && "contrast" %in% colnames(het_cv)) {
  filter(het_cv, contrast == !!contrast)
} else { NULL }
if (!is.null(het_ct) && nrow(het_ct) > 0 && !is.null(catalog) && nrow(catalog) > 0) {
  het_ct <- het_ct %>%
    left_join(select(catalog, gene_id, gene_name, class_code), by = "gene_id")
}

# Per-sample lncRNA detection from TPM matrix (QC)
tpm <- tryCatch(readRDS(params$tpm), error = function(e) NULL)
lncrna_per_sample <- NULL
if (!is.null(tpm) && !is.null(catalog) && nrow(catalog) > 0) {
  lncrna_ids <- catalog$gene_id[catalog$origin %in% c("novel", "annotated_lncrna")]
  lncrna_ids <- intersect(lncrna_ids, rownames(tpm))
  if (length(lncrna_ids) > 0) {
    tpm_lnc <- tpm[lncrna_ids, , drop = FALSE]
    lncrna_per_sample <- data.frame(
      sample        = colnames(tpm_lnc),
      n_novel       = colSums(tpm_lnc[rownames(tpm_lnc) %in% catalog$gene_id[catalog$origin == "novel"], , drop = FALSE] > 0.5),
      n_annotated   = colSums(tpm_lnc[rownames(tpm_lnc) %in% catalog$gene_id[catalog$origin == "annotated_lncrna"], , drop = FALSE] > 0.5),
      n_total_lnc   = colSums(tpm_lnc > 0.5),
      row.names     = NULL,
      stringsAsFactors = FALSE
    )
  }
}
```

## Tested catalog

```{r catalog-summary}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  missing_cat <- sum(is.na(de_ann$origin))
  if (missing_cat > 0.05 * nrow(de_ann)) {
    cat(sprintf("WARNING: %d / %d tested genes missing from gene catalog (possible gene_id drift between tx2gene and catalog).\n",
                missing_cat, nrow(de_ann)))
  }
  cat_sum <- de_ann %>%
    group_by(origin) %>%
    summarise(
      n_tested = n(),
      n_sig    = sum(!is.na(padj) & padj < 0.05),
      .groups  = "drop"
    ) %>%
    mutate(origin_label = ORIGIN_LABELS[origin]) %>%
    select(origin_label, n_tested, n_sig) %>%
    arrange(origin_label)
  kable(cat_sum, format = "html",
        col.names = c("Origin", "Genes tested", "Significant (padj<0.05)"))

  if (any(cat_sum$origin_label == ORIGIN_LABELS["reference_other"])) {
    cat("\n*\"Other reference genes (background)\" includes protein-coding and other non-lncRNA reference genes. ",
        "This category exists only so that DESeq2 normalizes over the full transcriptome and enrichment analyses ",
        "have a proper background — it is not a biologically interesting category for lncRNA-focused interpretation.*\n")
  }
} else {
  cat("No gene catalog available for this report.")
}
```

### lncRNA class codes {.unnumbered}

GFFCOMPARE assigns a `class_code` to each novel transcript. The code indicates the
relationship between the novel transcript and the reference annotation.

```{r class-code-legend}
legend_df <- data.frame(
  Code = names(CLASS_CODE_LEGEND),
  Description = unname(CLASS_CODE_LEGEND),
  stringsAsFactors = FALSE
)
kable(legend_df, format = "html", col.names = c("class_code", "Interpretation"))
```

```{r class-code-summary}
novel_de <- de_ann %>% filter(origin == "novel")
if (nrow(novel_de) > 0) {
  cc_sum <- novel_de %>%
    mutate(cc_simple = coalesce(class_code, "unknown")) %>%
    separate_rows(cc_simple, sep = ",\\s*") %>%
    filter(nzchar(cc_simple)) %>%
    group_by(class_code = cc_simple) %>%
    summarise(
      n_tested = n_distinct(gene_id),
      n_sig    = n_distinct(gene_id[!is.na(padj) & padj < 0.05]),
      .groups  = "drop"
    ) %>%
    arrange(desc(n_tested))
  cat(sprintf("Novel lncRNAs tested: %d  |  Significant: %d\n\n",
              nrow(novel_de),
              sum(novel_de$padj < 0.05, na.rm = TRUE)))
  kable(cc_sum, format = "html",
        col.names = c("class_code", "Tested", "Significant (padj<0.05)"))
} else {
  cat("No novel lncRNAs tested for this contrast.\n")
}
```

## lncRNA detection per sample (QC)

Number of lncRNAs (novel + annotated) detected per sample at TPM > 0.5.

```{r lncrna-per-sample, fig.width=8, fig.height=5}
if (!is.null(lncrna_per_sample) && nrow(lncrna_per_sample) > 0) {
  lps_long <- lncrna_per_sample %>%
    pivot_longer(c(n_novel, n_annotated, n_total_lnc),
                 names_to = "type", values_to = "n") %>%
    mutate(type = recode(type,
      n_novel       = "Novel",
      n_annotated   = "Annotated (reference)",
      n_total_lnc   = "Total lncRNA"
    ))
  p <- ggplot(lps_long, aes(reorder(sample, n), n, fill = type)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c(
      "Novel"                 = "#c2185b",
      "Annotated (reference)" = "#2e7d32",
      "Total lncRNA"          = "#1a237e"
    )) +
    labs(x = NULL, y = "N lncRNAs detected (TPM > 0.5)", fill = NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
} else {
  cat("TPM matrix not available for this report.\n")
}
```

## DE summary

```{r de-summary}
if (!is.null(de_ann) && nrow(de_ann) > 0) {
  n_sig <- sum(de_ann$padj < 0.05, na.rm = TRUE)
  n_up  <- sum(de_ann$padj < 0.05 & de_ann$log2FoldChange > 0, na.rm = TRUE)
  n_dn  <- sum(de_ann$padj < 0.05 & de_ann$log2FoldChange < 0, na.rm = TRUE)
  kable(data.frame(
    genes_tested = nrow(de_ann),
    significant_padj_0.05 = n_sig,
    up = n_up, down = n_dn
  ), format = "html")
} else {
  cat("No differential expression results found for this contrast.\n\n")
  cat("The contrast was declared in the analysis but no DE output was recorded upstream. ")
  cat("This usually means DESeq2 skipped or failed it (e.g. missing condition level or too few samples in a batch subset). ")
  cat("Check the pipeline log for `SKIP`/`ERROR` lines mentioning this contrast.\n")
}
```

### Volcano plot

Colored by `origin` (green: annotated lncRNAs, fuchsia: novel lncRNAs, gray: protein-coding/other).
Point size marks significance (large = padj<0.05, small = not significant); significant
genes are labeled.

```{r volcano, fig.width=8, fig.height=6}
if (!is.null(de_ann) && nrow(de_ann) > 0 &&
    all(c("log2FoldChange", "pvalue") %in% colnames(de_ann))) {
  de_ann <- de_ann %>% mutate(
    sig    = ifelse(!is.na(padj) & padj < 0.05, "FDR<0.05", "ns"),
    origin = ifelse(is.na(origin), "reference_other", origin),
    logp   = -log10(pvalue),
    label  = coalesce(gene_name, gene_id)
  )
  p <- ggplot(de_ann, aes(log2FoldChange, logp)) +
    geom_point(aes(color = origin, size = sig), alpha = 0.6) +
    geom_text_repel(
      data = filter(de_ann, sig == "FDR<0.05") %>%
               arrange(pvalue) %>% slice_head(n = 20),
      aes(label = label), size = 3, max.overlaps = Inf, min.segment.length = 0.2
    ) +
    scale_color_manual(
      values = ORIGIN_COLORS[sort(unique(de_ann$origin))],
      labels = ORIGIN_LABELS[sort(unique(de_ann$origin))],
      na.value = "#7f7f7f"
    ) +
    scale_size_manual(values = c("FDR<0.05" = 3, "ns" = 0.8)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.4) +
    labs(color = "Origin", size = "Significance") +
    theme_minimal()
  print(p)
} else {
  cat("No DE data for this contrast — volcano plot skipped.\n")
}
```

### Tested lncRNAs by class

Every tested class is shown, including those with zero hits. Light gray = not
significant, dark = significant (split by direction).

```{r class-barplot}
lnc_by_class <- NULL
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  lnc_by_class <- de_ann %>%
    filter(origin %in% c("novel", "annotated_lncrna"), !is.na(padj)) %>%
    mutate(
      label  = ifelse(origin == "novel",
                      sapply(coalesce(class_code, "unknown"), format_class_code),
                      coalesce(gene_biotype, "lncRNA")),
      origin_label = ORIGIN_LABELS[origin],
      status = case_when(
        padj >= 0.05        ~ "not significant",
        log2FoldChange > 0  ~ "significant (up)",
        TRUE                ~ "significant (down)"
      )
    ) %>%
    count(origin_label, label, status) %>%
    complete(nesting(origin_label, label), status, fill = list(n = 0))
}
if (!is.null(lnc_by_class) && nrow(lnc_by_class) > 0) {
  p <- ggplot(lnc_by_class, aes(label, n, fill = status)) +
    geom_col(position = "stack") +
    facet_wrap(~origin_label, scales = "free_x") +
    scale_fill_manual(values = c(
      "not significant"    = "#d9d9d9",
      "significant (up)"   = "#b71c1c",
      "significant (down)" = "#1565c0"
    )) +
    labs(x = NULL, y = "N genes", fill = NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
} else {
  cat("No lncRNA genes tested for this contrast.")
}
```

### DE tables

::: {.panel-tabset}

#### Overview — significant genes

```{r de-table-all}
if (!is.null(de_ann) && nrow(de_ann) > 0) {
  tab <- de_ann %>%
    filter(!is.na(padj) & padj < 0.05) %>%
    mutate(
      origin = recode(origin, !!!ORIGIN_LABELS),
      class_code_label = sapply(coalesce(class_code, ""), format_class_code)
    ) %>%
    select(any_of(c("gene_id", "gene_name", "origin", "gene_biotype", "class_code",
                    "class_code_label",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant DE genes in this contrast."
  }
} else {
  cat("No DE data.")
}
```

#### Annotated lncRNAs (reference)

```{r de-annotated}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  tab <- de_ann %>%
    filter(origin == "annotated_lncrna", !is.na(padj), padj < 0.05) %>%
    mutate(origin = recode(origin, !!!ORIGIN_LABELS)) %>%
    select(any_of(c("gene_id", "gene_name", "origin", "gene_biotype",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant annotated lncRNA genes in this contrast."
  }
} else { "Catalog origin not available." }
```

#### Novel lncRNAs

```{r de-novel}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  tab <- de_ann %>%
    filter(origin == "novel", !is.na(padj), padj < 0.05) %>%
    mutate(
      origin = recode(origin, !!!ORIGIN_LABELS),
      class_code_label = sapply(coalesce(class_code, ""), format_class_code)
    ) %>%
    select(any_of(c("gene_id", "class_code", "class_code_label",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant novel lncRNA genes in this contrast."
  }
} else {
  cat("Catalog origin not available.")
}
```

:::

## Cis-regulatory associations

lncRNAs often act on nearby genes. For every lncRNA (novel or annotated) this analysis
tests Spearman correlation between its expression and every protein-coding gene within
&plusmn;`r format(params$cis_window, big.mark = ",")` bp,
across all samples in this norm_group. Pairs below the FDR threshold are candidate
cis-regulatory targets (`target_gene_id`/`target_gene_name`) — a starting point for
follow-up, not proof of causality. `spearman_rho` gives the direction/strength of the
correlation and `padj` the BH-corrected significance.

```{r cis}
if (!is.null(cis_sig) && nrow(cis_sig) > 0) {
  cat("Significant cis pairs:\n")
  datatable(cis_sig, options = list(pageLength = 20))
} else {
  cat("No significant cis-regulatory associations found.\n")
}
```

## Functional enrichment — ORA

```{r ora-dotplot, fig.width=7, fig.height=5}
if (!is.null(ora) && nrow(ora) > 0 && all(c("GeneRatio", "p.adjust", "Count") %in% colnames(ora))) {
  top_ora <- ora %>%
    arrange(p.adjust) %>%
    slice_head(n = 10) %>%
    mutate(GeneRatioNum = sapply(strsplit(GeneRatio, "/"),
                                 function(x) as.numeric(x[1]) / as.numeric(x[2])))
  p <- ggplot(top_ora, aes(GeneRatioNum, reorder(Description, GeneRatioNum))) +
    geom_point(aes(size = Count, color = p.adjust)) +
    scale_color_gradient(low = "#b71c1c", high = "#1565c0") +
    labs(title = "Top 10 ORA terms", x = "Gene ratio", y = NULL,
         size = "Count", color = "p.adjust") +
    theme_minimal()
  print(p)
}
```

```{r ora}
if (!is.null(ora) && nrow(ora) > 0) {
  datatable(compact_list_columns(ora), escape = FALSE, options = list(pageLength = 20))
} else {
  cat("No ORA enrichment terms found for this contrast.\n")
  cat("The analysis completed but no gene-set passed the threshold.\n")
}
```

## Functional enrichment — GSEA

```{r gsea-ridgeplot, fig.width=7, fig.height=5}
if (!is.null(gsea) && nrow(gsea) > 0 && "core_enrichment" %in% colnames(gsea) &&
    !is.null(de_ct) && nrow(de_ct) > 0) {
  rank_col <- if ("stat" %in% colnames(de_ct)) "stat" else "log2FoldChange"
  rank_label <- if (rank_col == "stat") "DE test statistic" else "log2 fold-change"
  top_gsea <- gsea %>% arrange(p.adjust) %>% slice_head(n = 10)
  ridge_df <- top_gsea %>%
    select(Description, core_enrichment) %>%
    separate_rows(core_enrichment, sep = "/") %>%
    rename(gene_id = core_enrichment) %>%
    left_join(select(de_ct, gene_id, all_of(rank_col)), by = "gene_id") %>%
    filter(!is.na(.data[[rank_col]]))
  if (nrow(ridge_df) > 0) {
    p <- ggplot(ridge_df, aes(.data[[rank_col]], reorder(Description, .data[[rank_col]], FUN = median))) +
      geom_density_ridges(fill = "#6a1b9a", alpha = 0.6) +
      labs(title = "Top 10 GSEA terms", x = rank_label, y = NULL) +
      theme_minimal()
    print(p)
  } else {
    cat("Core-enrichment genes did not match the DE summary — ridgeplot skipped.\n")
  }
}
```

```{r gsea}
if (!is.null(gsea) && nrow(gsea) > 0) {
  datatable(compact_list_columns(gsea), escape = FALSE, options = list(pageLength = 20))
} else {
  cat("No GSEA enrichment found for this contrast.\n")
  cat("The ranked gene-set analysis completed with no significant sets.\n")
}
```

```{r heterogeneity-check, include=FALSE}
# Off by default: het_ct/het_cv_ct are NULL (or 0 rows) unless
# params.heterogeneity_analysis is enabled for this norm_group, in which case this
# whole section is skipped and the report is identical to the pre-feature pipeline.
show_heterogeneity <- !is.null(het_ct) && nrow(het_ct) > 0
```

`r if (show_heterogeneity) "## Heterogeneity Analysis"`

```{r heterogeneity-intro, results='asis'}
if (show_heterogeneity) {
  cat('<div style="text-align: justify;">\n\n')
  cat("Complementary, patient-level diagnostic (Carrasco-Leon et al. 2021, *Leukemia*): ",
      "a standard group-mean DE test can miss lncRNAs whose expression is heterogeneous ",
      "across individual case samples. This does not replace the DESeq2 result above.\n\n")
  cat('</div>\n')
}
```

```{r heterogeneity-cv-plot, eval=show_heterogeneity, fig.width=6, fig.height=4, results='asis'}
if (!is.null(het_cv_ct) && nrow(het_cv_ct) > 0) {
  p <- ggplot(het_cv_ct, aes(origin_group, cv, fill = origin_group)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    labs(x = NULL, y = "CV (log2 TPM, case samples)", fill = NULL) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)

  t_stat <- unique(het_cv_ct$t_statistic)[1]
  p_val  <- unique(het_cv_ct$p_value)[1]
  cat(sprintf(
    "\n- lncRNA CV: mean = %.3f (n = %d)\n- Coding CV: mean = %.3f (n = %d)\n- Welch t-test: t = %.3f, p = %.3g\n",
    unique(het_cv_ct$mean_cv_lncrna)[1], unique(het_cv_ct$n_lncrna_genes)[1],
    unique(het_cv_ct$mean_cv_coding)[1], unique(het_cv_ct$n_coding_genes)[1],
    ifelse(is.na(t_stat), NA, t_stat), ifelse(is.na(p_val), NA, p_val)
  ))
} else {
  cat("No CV comparison data available for this contrast.\n")
}
```

### Altered lncRNAs missed by the standard DE test

Genes called "altered" by sample-level concordance (concordant in
&ge;`r round(100 * params$heterogeneity_concordant_frac)`% of case samples,
discordant in <`r round(100 * params$heterogeneity_discordant_frac)`%) that
were **not** already significant (padj < 0.05) in the main DESeq2 contrast — the actual
value-add of this diagnostic.

```{r heterogeneity-table}
if (show_heterogeneity) {
  de_sig_ids <- if (!is.null(de_ct) && "gene_id" %in% colnames(de_ct)) {
    de_ct$gene_id[!is.na(de_ct$padj) & de_ct$padj < 0.05]
  } else character()

  tab <- het_ct %>%
    filter(altered_call == "altered", !(gene_id %in% de_sig_ids)) %>%
    select(any_of(c("gene_id", "gene_name", "class_code", "direction",
                    "concordant_frac", "discordant_frac", "n_case", "n_control")))
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    cat("No additional altered lncRNAs found beyond the standard DESeq2 result.\n")
  }
}
```

## Conservation

```{r conservation}
if (has_conservation) {
  cat(sprintf("Metric: %s (build: %s)\n\n",
              params$conservation_metric, params$genome_build))
  # Scoped to genes tested in this contrast (de_ann), not the whole catalog —
  # otherwise every report would show the same genome-wide summary.
  cons_data <- de_ann %>% filter(!is.na(conservation_mean))
  cons_sum <- cons_data %>%
    group_by(origin) %>%
    summarise(
      n           = n(),
      mean_score  = round(mean(conservation_mean, na.rm = TRUE), 3),
      pct_conserved = round(100 * mean(conservation_pct, na.rm = TRUE), 1),
      .groups     = "drop"
    )
  if (nrow(cons_sum) > 0) {
    kable(cons_sum, format = "html")
  } else {
    cat("No tested genes with conservation data for this contrast. ")
    if (sum(de_ann$origin == "novel", na.rm = TRUE) == 0) {
      cat("This is expected when no novel lncRNAs passed the DESeq2 independent filtering step ")
      cat("for this contrast (e.g. low-abundance novel transcripts in primary samples with shallow sequencing). ")
    }
    cat("Conservation scores are only computed for novel lncRNAs.\n")
  }
} else {
  cat("Conservation not evaluated for this run (conservation_bigwig not configured).\n")
}
```

### Conservation vs differential expression

Splits the same genes by DE outcome in this contrast — is conservation higher or
lower among the ones that actually changed?

```{r conservation-by-sig}
if (has_conservation && exists("cons_data") && nrow(cons_data) > 0 && "padj" %in% colnames(cons_data)) {
  cons_by_sig <- cons_data %>%
    filter(!is.na(padj)) %>%
    mutate(sig = ifelse(padj < 0.05, "significant (padj<0.05)", "not significant")) %>%
    group_by(origin, sig) %>%
    summarise(
      n             = n(),
      mean_score    = round(mean(conservation_mean, na.rm = TRUE), 3),
      pct_conserved = round(100 * mean(conservation_pct, na.rm = TRUE), 1),
      .groups       = "drop"
    )
  if (nrow(cons_by_sig) > 0) {
    kable(cons_by_sig, format = "html")
  }
}
```

```{r conservation-plot, fig.width=7, fig.height=4}
if (has_conservation && exists("cons_data") && nrow(cons_data) > 0 && "padj" %in% colnames(cons_data)) {
  plot_df <- cons_data %>%
    filter(!is.na(padj)) %>%
    mutate(sig = ifelse(padj < 0.05, "significant", "not significant"))
  if (nrow(plot_df) > 0) {
    p <- ggplot(plot_df, aes(sig, conservation_mean, fill = sig)) +
      geom_violin(alpha = 0.6) +
      geom_boxplot(width = 0.15, outlier.shape = NA) +
      facet_wrap(~origin, scales = "free_y") +
      labs(x = NULL, y = "Mean phastCons score", fill = NULL) +
      theme_minimal() +
      theme(legend.position = "none")
    print(p)
  }
}
```

### Most conserved significant novel lncRNAs

```{r conservation-top-table}
if (has_conservation && exists("cons_data")) {
  top_cons <- cons_data %>%
    filter(origin == "novel", !is.na(padj), padj < 0.05) %>%
    select(any_of(c("gene_id", "class_code", "conservation_mean", "conservation_pct",
                    "log2FoldChange", "padj"))) %>%
    arrange(desc(conservation_mean))
  if (nrow(top_cons) > 0) {
    datatable(top_cons, filter = "top", options = list(pageLength = 10))
  } else {
    cat("No significant novel lncRNAs with conservation data in this contrast.\n")
  }
}
```

## Synteny

```{r synteny}
if (has_synteny) {
  cat(sprintf("Target species: %s\n\n", params$synteny_target_species))
  # Scoped to genes tested in this contrast (de_ann), not the whole catalog —
  # otherwise every report would show the same genome-wide summary.
  syn_data <- de_ann %>% filter(!is.na(syntenic_locus))
  syn_sum <- syn_data %>%
    group_by(origin) %>%
    summarise(
      n             = n(),
      n_syntenic    = sum(syntenic_locus == TRUE, na.rm = TRUE),
      pct_syntenic  = round(100 * mean(syntenic_locus == TRUE, na.rm = TRUE), 1),
      .groups       = "drop"
    )
  if (nrow(syn_sum) > 0) {
    kable(syn_sum, format = "html")
  } else {
    cat("No tested genes with synteny data for this contrast. ")
    if (sum(de_ann$origin == "novel", na.rm = TRUE) == 0) {
      cat("This is expected when no novel lncRNAs passed the DESeq2 independent filtering step ")
      cat("for this contrast (e.g. low-abundance novel transcripts in primary samples with shallow sequencing). ")
    }
    cat("Synteny checks are only computed for novel lncRNAs.\n")
  }
} else {
  cat("Synteny not evaluated for this run (synteny_chain_file/synteny_target_gtf not configured).\n")
}
```

### Synteny vs differential expression

```{r synteny-by-sig}
if (has_synteny && exists("syn_data") && nrow(syn_data) > 0 && "padj" %in% colnames(syn_data)) {
  syn_by_sig <- syn_data %>%
    filter(!is.na(padj)) %>%
    mutate(sig = ifelse(padj < 0.05, "significant (padj<0.05)", "not significant")) %>%
    group_by(origin, sig) %>%
    summarise(
      n             = n(),
      n_syntenic    = sum(syntenic_locus == TRUE, na.rm = TRUE),
      pct_syntenic  = round(100 * mean(syntenic_locus == TRUE, na.rm = TRUE), 1),
      .groups       = "drop"
    )
  if (nrow(syn_by_sig) > 0) {
    kable(syn_by_sig, format = "html")
  }
}
```

### Significant novel lncRNAs with a syntenic locus

Candidate orthologous loci in `r params$synteny_target_species` — a starting point for
cross-species follow-up, not proof of conserved function.

```{r synteny-top-table}
if (has_synteny && exists("syn_data")) {
  top_syn <- syn_data %>%
    filter(origin == "novel", syntenic_locus == TRUE, !is.na(padj), padj < 0.05) %>%
    select(any_of(c("gene_id", "class_code", "syntenic_target_gene_id",
                    "log2FoldChange", "padj")))
  if (nrow(top_syn) > 0) {
    datatable(top_syn, filter = "top", options = list(pageLength = 10))
  } else {
    cat("No significant novel lncRNAs with a syntenic locus in this contrast.\n")
  }
}
```

## Session info

```{r session, echo=FALSE}
cat("Generated:", format(Sys.time()), "\n")
si <- tryCatch({
  if (requireNamespace("sessioninfo", quietly = TRUE)) sessioninfo::session_info() else sessionInfo()
}, error = function(e) NULL)
if (!is.null(si)) print(si)
```