---
title: "lncRNA report"
subtitle: "norm_group report"
format:
  html:
    self-contained: true
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
params:
  contrast: ""
  group: ""
  de_summary: null
  cis_pairs: null
  cis_pairs_sig: null
  cis_window: 100000
  ora: null
  gsea: null
  heterogeneity: null
  heterogeneity_cv: null
  heterogeneity_concordant_frac: 0.50
  heterogeneity_discordant_frac: 0.25
  gene_catalog: null
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(knitr)
library(DT)
library(plotly)
library(ggplot2)
library(ggrepel)
library(ggridges)

contrast <- params$contrast
group    <- params$group

ORIGIN_COLORS <- c(
  "novel"           = "#c2185b",
  "annotated_lncrna" = "#2e7d32",
  "reference_other" = "#7f7f7f"
)

# Compact long gene-list columns into "gene1, gene2, ... (+N more)" cells
# with the full list available on hover (title attribute).
compact_gene_col <- function(x, max_show = 3) {
  vapply(x, function(s) {
    if (is.na(s) || nchar(s) == 0) return(NA_character_)
    genes <- strsplit(s, ",\\s*", perl = TRUE)[[1]]
    genes <- genes[nzchar(genes)]
    n <- length(genes)
    if (n == 0) return("")
    shown <- paste(genes[seq_len(min(n, max_show))], collapse = ", ")
    suffix <- if (n > max_show) sprintf(" (+%d more)", n - max_show) else ""
    full <- paste(genes, collapse = ", ")
    sprintf('<span title="%s">%s%s</span>',
            htmltools::htmlEscape(full), htmltools::htmlEscape(shown), suffix)
  }, character(1))
}
# Mark gene-list columns for compacting; escape every other character column
# so the table can be rendered with escape = FALSE.
compact_list_columns <- function(tab) {
  for (nm in names(tab)) {
    x <- tab[[nm]]
    if (!is.character(x)) next
    if (any(nchar(x[!is.na(x)]) > 40)) {
      tab[[nm]] <- compact_gene_col(x)
    } else {
      tab[[nm]] <- htmltools::htmlEscape(x)
    }
  }
  tab
}
```

## Contrast: `r contrast`

```{r load, include=FALSE}
de <- tryCatch(
  read_tsv(params$de_summary, show_col_types = FALSE),
  error = function(e) NULL
)
de_ct <- if (!is.null(de) && "contrast" %in% colnames(de)) {
  filter(de, contrast == !!contrast)
} else { de }

catalog <- tryCatch(read_tsv(params$gene_catalog, show_col_types = FALSE),
                    error = function(e) NULL)

de_ann <- de_ct
if (!is.null(catalog) && nrow(catalog) > 0 && "gene_id" %in% colnames(de_ct)) {
  de_ann <- de_ct %>%
    left_join(select(catalog, gene_id, gene_name, gene_biotype, origin, class_code),
              by = "gene_id")
}

cis <- tryCatch(read_tsv(params$cis_pairs, show_col_types = FALSE),
                error = function(e) NULL)
cis_sig <- tryCatch(read_tsv(params$cis_pairs_sig, show_col_types = FALSE),
                    error = function(e) NULL)
ora <- tryCatch(read_tsv(params$ora, show_col_types = FALSE),
                error = function(e) NULL)
gsea <- tryCatch(read_tsv(params$gsea, show_col_types = FALSE),
                 error = function(e) NULL)

# Restrict enrichment tables to the current contrast
if (!is.null(ora) && "contrast" %in% colnames(ora)) {
  ora <- filter(ora, contrast == !!contrast)
}
if (!is.null(gsea) && "contrast" %in% colnames(gsea)) {
  gsea <- filter(gsea, contrast == !!contrast)
}

# Heterogeneity diagnostic (optional feature, off by default): the group-level table is
# a header-only, 0-row placeholder when params.heterogeneity_analysis is disabled for
# this norm_group, so het_ct/het_cv_ct simply end up empty and the section is skipped.
het <- tryCatch(read_tsv(params$heterogeneity, show_col_types = FALSE),
                error = function(e) NULL)
het_cv <- tryCatch(read_tsv(params$heterogeneity_cv, show_col_types = FALSE),
                   error = function(e) NULL)
het_ct <- if (!is.null(het) && "contrast" %in% colnames(het)) {
  filter(het, contrast == !!contrast)
} else { NULL }
het_cv_ct <- if (!is.null(het_cv) && "contrast" %in% colnames(het_cv)) {
  filter(het_cv, contrast == !!contrast)
} else { NULL }
if (!is.null(het_ct) && nrow(het_ct) > 0 && !is.null(catalog) && nrow(catalog) > 0) {
  het_ct <- het_ct %>%
    left_join(select(catalog, gene_id, gene_name, class_code), by = "gene_id")
}
```

## Tested catalog

```{r catalog-summary}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  missing_cat <- sum(is.na(de_ann$origin))
  if (missing_cat > 0.05 * nrow(de_ann)) {
    cat(sprintf("WARNING: %d / %d tested genes missing from gene catalog (possible gene_id drift between tx2gene and catalog).\n",
                missing_cat, nrow(de_ann)))
  }
  cat_sum <- de_ann %>%
    group_by(origin) %>%
    summarise(
      n_tested = n(),
      n_sig    = sum(!is.na(padj) & padj < 0.05),
      .groups  = "drop"
    ) %>%
    arrange(origin)
  kable(cat_sum, format = "html", col.names = c("Origin", "Genes tested", "Significant (padj<0.05)"))
} else {
  cat("No gene catalog available for this report.")
}
```

## DE summary

```{r de-summary}
if (!is.null(de_ann) && nrow(de_ann) > 0) {
  n_sig <- sum(de_ann$padj < 0.05, na.rm = TRUE)
  n_up  <- sum(de_ann$padj < 0.05 & de_ann$log2FoldChange > 0, na.rm = TRUE)
  n_dn  <- sum(de_ann$padj < 0.05 & de_ann$log2FoldChange < 0, na.rm = TRUE)
  kable(data.frame(
    genes_tested = nrow(de_ann),
    significant_padj_0.05 = n_sig,
    up = n_up, down = n_dn
  ), format = "html")
} else {
  cat("**No differential expression results found for this contrast.**\n\n")
  cat("The contrast was declared in the analysis but no DE output was recorded upstream. ")
  cat("This usually means DESeq2 skipped or failed it (e.g. missing condition level or too few samples in a batch subset). ")
  cat("Check the pipeline log for `SKIP`/`ERROR` lines mentioning this contrast.\n")
}
```

### Volcano plot

Colored by `origin` (green: annotated lncRNAs, fuchsia: novel lncRNAs, gray: protein-coding/other).
Point size marks significance (large = padj<0.05, small = not significant); significant
genes are labeled.

```{r volcano, fig.width=8, fig.height=6}
if (!is.null(de_ann) && nrow(de_ann) > 0 &&
    all(c("log2FoldChange", "pvalue") %in% colnames(de_ann))) {
  de_ann <- de_ann %>% mutate(
    sig    = ifelse(!is.na(padj) & padj < 0.05, "FDR<0.05", "ns"),
    origin = ifelse(is.na(origin), "reference_other", origin),
    logp   = -log10(pvalue),
    label  = coalesce(gene_name, gene_id)
  )
  p <- ggplot(de_ann, aes(log2FoldChange, logp)) +
    geom_point(aes(color = origin, size = sig), alpha = 0.6) +
    geom_text_repel(
      data = filter(de_ann, sig == "FDR<0.05"),
      aes(label = label), size = 3, max.overlaps = 30, min.segment.length = 0
    ) +
    scale_color_manual(values = ORIGIN_COLORS[sort(unique(de_ann$origin))],
                       na.value = "#7f7f7f") +
    scale_size_manual(values = c("FDR<0.05" = 3, "ns" = 0.8)) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.4) +
    labs(color = "Origin", size = "Significance") +
    theme_minimal()
  print(p)
} else {
  cat("No DE data for this contrast — volcano plot skipped.\n")
}
```

### Tested lncRNAs by class

Every tested class is shown, including those with zero hits. Light gray = not
significant, dark = significant (split by direction).

```{r class-barplot}
lnc_by_class <- NULL
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  lnc_by_class <- de_ann %>%
    filter(origin %in% c("novel", "annotated_lncrna"), !is.na(padj)) %>%
    mutate(
      label  = ifelse(origin == "novel",
                      coalesce(class_code, "unknown"),
                      coalesce(gene_biotype, "lncRNA")),
      status = case_when(
        padj >= 0.05        ~ "not significant",
        log2FoldChange > 0  ~ "significant (up)",
        TRUE                ~ "significant (down)"
      )
    ) %>%
    count(origin, label, status) %>%
    complete(nesting(origin, label), status, fill = list(n = 0))
}
if (!is.null(lnc_by_class) && nrow(lnc_by_class) > 0) {
  p <- ggplot(lnc_by_class, aes(label, n, fill = status)) +
    geom_col(position = "stack") +
    facet_wrap(~origin, scales = "free_x") +
    scale_fill_manual(values = c(
      "not significant"    = "#d9d9d9",
      "significant (up)"   = "#b71c1c",
      "significant (down)" = "#1565c0"
    )) +
    labs(x = NULL, y = "N genes", fill = NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
} else {
  cat("No lncRNA genes tested for this contrast.")
}
```

### DE tables

::: {.panel-tabset}

#### Overview — significant genes

```{r de-table-all}
if (!is.null(de_ann) && nrow(de_ann) > 0) {
  tab <- de_ann %>%
    filter(!is.na(padj) & padj < 0.05) %>%
    select(any_of(c("gene_id", "gene_name", "origin", "gene_biotype", "class_code",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant DE genes in this contrast."
  }
} else {
  cat("No DE data.")
}
```

#### Annotated lncRNAs

```{r de-annotated}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  tab <- de_ann %>%
    filter(origin == "annotated_lncrna", !is.na(padj), padj < 0.05) %>%
    select(any_of(c("gene_id", "gene_name", "gene_biotype",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant annotated lncRNA genes in this contrast."
  }
} else { "Catalog origin not available." }
```

#### Novel lncRNAs

```{r de-novel}
if (!is.null(de_ann) && nrow(de_ann) > 0 && "origin" %in% colnames(de_ann)) {
  tab <- de_ann %>%
    filter(origin == "novel", !is.na(padj), padj < 0.05) %>%
    select(any_of(c("gene_id", "class_code",
                    "baseMean", "log2FoldChange", "lfcSE", "pvalue", "padj"))) %>%
    arrange(padj)
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    "No significant novel lncRNA genes in this contrast."
  }
} else {
  cat("Catalog origin not available.")
}
```

:::

## Cis-regulatory associations

lncRNAs often act on nearby genes. For every lncRNA (novel or annotated) this analysis
tests Spearman correlation between its expression and every protein-coding gene within
&plusmn;`r format(params$cis_window, big.mark = ",")` bp,
across all samples in this norm_group. Pairs below the FDR threshold are candidate
cis-regulatory targets (`target_gene_id`/`target_gene_name`) — a starting point for
follow-up, not proof of causality. `spearman_rho` gives the direction/strength of the
correlation and `padj` the BH-corrected significance.

```{r cis}
if (!is.null(cis_sig) && nrow(cis_sig) > 0) {
  cat("**Significant cis pairs:**\n")
  datatable(cis_sig, options = list(pageLength = 20))
} else {
  cat("**No significant cis-regulatory associations found.**\n")
}
```

## Functional enrichment — ORA

```{r ora-dotplot, fig.width=7, fig.height=5}
if (!is.null(ora) && nrow(ora) > 0 && all(c("GeneRatio", "p.adjust", "Count") %in% colnames(ora))) {
  top_ora <- ora %>%
    arrange(p.adjust) %>%
    slice_head(n = 10) %>%
    mutate(GeneRatioNum = sapply(strsplit(GeneRatio, "/"),
                                 function(x) as.numeric(x[1]) / as.numeric(x[2])))
  p <- ggplot(top_ora, aes(GeneRatioNum, reorder(Description, GeneRatioNum))) +
    geom_point(aes(size = Count, color = p.adjust)) +
    scale_color_gradient(low = "#b71c1c", high = "#1565c0") +
    labs(title = "Top 10 ORA terms", x = "Gene ratio", y = NULL,
         size = "Count", color = "p.adjust") +
    theme_minimal()
  print(p)
}
```

```{r ora}
if (!is.null(ora) && nrow(ora) > 0) {
  datatable(compact_list_columns(ora), escape = FALSE, options = list(pageLength = 20))
} else {
  cat("**No ORA enrichment terms found for this contrast.**\n")
  cat("The analysis completed but no gene-set passed the threshold.\n")
}
```

## Functional enrichment — GSEA

```{r gsea-ridgeplot, fig.width=7, fig.height=5}
if (!is.null(gsea) && nrow(gsea) > 0 && "core_enrichment" %in% colnames(gsea) &&
    !is.null(de_ct) && "stat" %in% colnames(de_ct)) {
  top_gsea <- gsea %>% arrange(p.adjust) %>% slice_head(n = 10)
  ridge_df <- top_gsea %>%
    select(Description, core_enrichment) %>%
    separate_rows(core_enrichment, sep = "/") %>%
    rename(gene_id = core_enrichment) %>%
    left_join(select(de_ct, gene_id, stat), by = "gene_id") %>%
    filter(!is.na(stat))
  if (nrow(ridge_df) > 0) {
    p <- ggplot(ridge_df, aes(stat, reorder(Description, stat, FUN = median))) +
      geom_density_ridges(fill = "#6a1b9a", alpha = 0.6) +
      labs(title = "Top 10 GSEA terms", x = "DE test statistic", y = NULL) +
      theme_minimal()
    print(p)
  } else {
    cat("Core-enrichment genes did not match the DE summary — ridgeplot skipped.\n")
  }
}
```

```{r gsea}
if (!is.null(gsea) && nrow(gsea) > 0) {
  datatable(compact_list_columns(gsea), escape = FALSE, options = list(pageLength = 20))
} else {
  cat("**No GSEA enrichment found for this contrast.**\n")
  cat("The ranked gene-set analysis completed with no significant sets.\n")
}
```

```{r heterogeneity-check, include=FALSE}
# Off by default: het_ct/het_cv_ct are NULL (or 0 rows) unless
# params.heterogeneity_analysis is enabled for this norm_group, in which case this
# whole section is skipped and the report is identical to the pre-feature pipeline.
show_heterogeneity <- !is.null(het_ct) && nrow(het_ct) > 0
```

`r if (show_heterogeneity) "## Heterogeneity Analysis"`

```{r heterogeneity-intro, results='asis'}
if (show_heterogeneity) {
  cat('<div style="text-align: justify;">\n\n')
  cat("Complementary, patient-level diagnostic (Carrasco-Leon et al. 2021, *Leukemia*): ",
      "a standard group-mean DE test can miss lncRNAs whose expression is heterogeneous ",
      "across individual case samples. This does not replace the DESeq2 result above.\n\n")
  cat('</div>\n')
}
```

```{r heterogeneity-cv-plot, eval=show_heterogeneity, fig.width=6, fig.height=4, results='asis'}
if (!is.null(het_cv_ct) && nrow(het_cv_ct) > 0) {
  p <- ggplot(het_cv_ct, aes(origin_group, cv, fill = origin_group)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.15, outlier.shape = NA) +
    labs(x = NULL, y = "CV (log2 TPM, case samples)", fill = NULL) +
    theme_minimal() +
    theme(legend.position = "none")
  print(p)

  t_stat <- unique(het_cv_ct$t_statistic)[1]
  p_val  <- unique(het_cv_ct$p_value)[1]
  cat(sprintf(
    "\n- **lncRNA CV**: mean = %.3f (n = %d)\n- **Coding CV**: mean = %.3f (n = %d)\n- **Welch t-test**: t = %.3f, p = %.3g\n",
    unique(het_cv_ct$mean_cv_lncrna)[1], unique(het_cv_ct$n_lncrna_genes)[1],
    unique(het_cv_ct$mean_cv_coding)[1], unique(het_cv_ct$n_coding_genes)[1],
    ifelse(is.na(t_stat), NA, t_stat), ifelse(is.na(p_val), NA, p_val)
  ))
} else {
  cat("No CV comparison data available for this contrast.\n")
}
```

### Altered lncRNAs missed by the standard DE test

Genes called "altered" by sample-level concordance (concordant in
&ge;`r round(100 * params$heterogeneity_concordant_frac)`% of case samples,
discordant in <`r round(100 * params$heterogeneity_discordant_frac)`%) that
were **not** already significant (padj < 0.05) in the main DESeq2 contrast — the actual
value-add of this diagnostic.

```{r heterogeneity-table}
if (show_heterogeneity) {
  de_sig_ids <- if (!is.null(de_ct) && "gene_id" %in% colnames(de_ct)) {
    de_ct$gene_id[!is.na(de_ct$padj) & de_ct$padj < 0.05]
  } else character()

  tab <- het_ct %>%
    filter(altered_call == "altered", !(gene_id %in% de_sig_ids)) %>%
    select(any_of(c("gene_id", "gene_name", "class_code", "direction",
                    "concordant_frac", "discordant_frac", "n_case", "n_control")))
  if (nrow(tab) > 0) {
    datatable(tab, filter = "top", options = list(pageLength = 20))
  } else {
    cat("No additional altered lncRNAs found beyond the standard DESeq2 result.\n")
  }
}
```

## Session info

```{r session, echo=FALSE}
cat("Generated:", format(Sys.time()), "\n")
si <- tryCatch({
  if (requireNamespace("sessioninfo", quietly = TRUE)) sessioninfo::session_info() else sessionInfo()
}, error = function(e) NULL)
if (!is.null(si)) print(si)
```